---
title: "Introducing discretefit"
output: rmarkdown::html_vignette
author: Josh McCormick
date: January 24, 2022
vignette: >
  %\VignetteIndexEntry{package_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package `discretefit` implements Monte Carlo simulations for goodness-of-fit (GOF) tests for discrete distributions. This includes tests based on the root-mean-square statistic, the Chi-squared statistic, the log-likelihood-ratio ($G^2$) statistic, the Freeman-Tukey (Hellinger-distance) statistic, the Kolmogorov-Smirnov statistic, and the Cramer-von Mises statistic.

Simulations are written in C++ (utilizing `Rcpp`) and are considerably faster than the simulated Chi-squared GOF test in the R `stats` package. 

## Usage

The GOF tests in `discretefit` function on a vector of counts, x, and a vector of probabilities, p. In the below example, x represents a vector of counts for five categories, and p represents a vector of probabilities for each corresponding category. The GOF tests provides p-values for the null hypothesis that x is a random sample of the discrete distribution defined by p. 

```{r setup, warning = FALSE}
library(discretefit)
library(bench)

x <- c(42, 0, 13, 2, 109)
p <- c(0.2, 0.05, 0.1, 0.05, 0.6)

chisq_gof(x, p)
rms_gof(x, p)
g_gof(x, p)
ft_gof(x, p)
ks_gof(x, p)
cvm_gof(x, p)
```

## Root-mean-square statistic

In a surprising number of cases, a simulated GOF test based on the root-mean-square statistic outperforms the Chi-squared test and other tests in the Cressie-Read power divergence family. This has been demonstrated by Perkins, Tygert, and Ward (2011). They provide the following toy example.

Take a discrete distribution with 50 bins (or categories). The probability for the first and second bin is 0.25. The probability for each of the remaining 48 bins is 0.5 / 48 (~0.0104).  

Now take the observed counts of 15 for the first bin, 5 for the second bin, and zero for each of the remaining 48 bins. It's obvious that these observations are very unlikely to occur for a random sample from the above distribution. However, the Chi-squared, $G^2$, and Freeman-Tukey tests fail to reject the null hypothesis. 

```{r}
x <- c(15, 5, rep(0, 48))
p <- c(0.25, 0.25, rep(1/(2 * 50 -4), 48))

chisq_gof(x, p)
g_gof(x, p)
ft_gof(x, p)
```

By contrast, the root-mean-square test convincingly rejects the null hypothesis.

```{r}
rms_gof(x, p)
```

For additional examples, also see Perkins, Tygert, and Ward (2011) and Ward and Carroll (2014). 

## Speed

The simulated Chi-squared GOF test in `discretefit` produces identical p-values to the simulated Chi-squared GOF test in the `stats` package that is part of base R. 

```{r}
set.seed(499)
chisq_gof(x, p, reps = 2000)$p.value
set.seed(499)
chisq.test(x, p = p, simulate.p.value = TRUE)$p.value
```

However, because Monte Carlo simulations in `discretefit` are implemented in C++, `chisq_gof` is much faster than `chisq.test`, especially when a large number of simulations are required.

```{r}
bench::system_time(
  chisq_gof(x, p, reps = 20000)
)

bench::system_time(
  chisq.test(x, p = p, simulate.p.value = TRUE, B = 20000)
)
```

Additionally, the simulated GOF tests in base R is vectorized, so for large vectors attempting a large number of simulations may not be possible because of memory constraints. Since the functions in `discretefit` are not vectorized, memory use is minimized. 

## Computation

Let $N$ represent the number of observations, let $k$ represent the number of bins, let $p_{i}$ represent the expected fraction of observations in the $i^{th}$ bin, and let $q_{i}$ represent the observed fraction of observations in the $i^{th}$ bin. Pearson's chi-squared is defined as follows (Pearson, 1900):

$$
X^2 = \sum_{i = 1}^{k} \frac{(q_{i} - p_{i})^2} {p_{i}} 
$$
The log-likelihood ratio statistic is defined as follows under the convention that $q_{i}$ ln($\frac{q_{i}} {p_{i}}$) = 0 when $q_{i}$ = 0.

$$ 
G^2 = 2N \sum_{i = 1}^{k} q_{i} \ln(\frac{q_{i}} {p_{i}}) 
$$
The Freeman-Tukey statistic was introduced by Freeman and Tukey (1950). For application to GOF tests see Cressie and Read (1984).

$$ FT = 4N \sum_{i = 1}^{k} (\sqrt{q_{i}} - \sqrt{p_{i}})^2 $$
Finally, the root-mean-square statistic is defined as follows:

$$ RMS = \sqrt{N^{-1} \sum_{i = 1}^{k} (q_{i} - p_{i})^2}$$

For statistics based on the empirical cumulative distribution function, let $S_{i}$ correspond to the empirical cumulative distribution function for the observed data, $T_{i}$ correspond to the cumulative distribution function for the theoretical distribution, and $Z_{i}$ = $S_{i}$ - $T_{i}$.

The Kolmogorov-Smirnov statistics for discrete distributions as defined by Pettitt and Stephens is as follows (1977):

$$
KS = \max_{1 \le i \le k} |Z_i|
$$

The `cvm_gof` function utilizes the formula for the Cramer-von Mises statistic for discrete distributions originally introduced by Choulakian, Lockhart, and Stephens (1994). 

$$ CVM =  N^{-1} \sum_{i = 1}^{k} Z_{i}^2p_{i} $$
For an alternative formula see Lockhart, Spinelli, and Stephens (2007).

All p-values calculated by the functions in `discretefit` follow the formula for simulated p-values proposed by Dwass (1957). For the below equation, let m represent the number of simulations and let B represent the number of simulations where the test statistic for the simulated data is greater than or equal to the test statistic for the observed data. 

$${p}_{u} = P(B <= b) = \frac{b + 1}{m + 1}$$

This is the equation used to calculate simulated p-values in the `chisq.test` function in the `stats` package but some implementations of simulated p-values, for example the `dgof` package, use the unbiased estimator, $\frac{B}{m}$. For an explanation of why the biased estimator yields a test of the correct size and the unbiased estimator does not, see Phipson and Smyth (2011). 

## Alternatives 

Several other packages implement GOF tests for discrete distributions.

As noted above, the `stats` package in base R implements a simulated  Chi-squared GOF test.

I'm not aware of an R package that implements a simulated $G^2$ GOF test but the packages `RVAideMemoire` and `DescTools` implement GOF tests that utilize approximations based on the Chi-squared distribution.

The `dgof` package (Anderson and Emerson 2011) implements simulated Kolmogorov-Smirnov GOF tests and simulated Cramer-von Mises GOF tests . The `cvmdisc` package also implements a simulated Cramer-von Mises GOF test. 

The `KSgeneral` package (Dimitrova, Kaishev, and Tan, 2020) implements exact Kolmogorov-Smirnov GOF tests and fast, simulated GOF tests utilizing the algorithm introduced by Wood and Altavela (1978).

I'm not aware of another R package that implements a root-mean-square GOF test.

## References

Arnold, Taylor B. and John W. Emerson. “Nonparametric goodness-of-fit tests for discrete null distributions.” R Journal. https://doi.org/10.32614/rj-2011-016 

Choulakian, V., Richard. A. Lockhart, and Michael A. Stephens. "Cramer–von Mises statistics for discrete distributions." Canadian Journal of Statistics, 1994. https://doi.org/10.2307/3315828 

Cressie, N. and Read, T.R. "Multinomial goodness-of-fit tests." Journal of the Royal Statistical Society, 1984. https://doi.org/10.1111/j.2517-6161.1984.tb01318.x

Dimitrova, Dimitrina S., Vladimir K. Kaishev, and Senren Tan. "Computing the Kolmogorov-Smirnov distribution when the underlying CDF is purely discrete, mixed, or continuous." Journal of Statistical Software, 2020. https://doi.org/10.18637/jss.v095.i10 

Dwass, Meyer. "Modified randomization tests for nonparametric hypotheses." Annuls of Mathematical Statistics, 1957.  https://doi.org/10.1214/aoms/1177707045  

Eddelbuettel, Dirk and Romain Francois. "Rcpp: Seamless R and C++ Integration." Journal of Statistical Software, 2011. https://www.jstatsoft.org/article/view/v040i08

Freeman, Murray F. and John W. Tukey. "Transformations related to the angular and the square root." The Annals of Mathematical Statistics, 1950. https://doi.org/10.1214/aoms/1177729756

Lockhart, Richard A., John J. Spinelli and Michael A. Stephens. "Cramer–von Mises statistics for discrete distributions with unknown parameters." Canadian Journal of Statistics, 2007. https://doi.org/10.1002/cjs.5550350111 

Pearson, Karl. "On the criterion that a given system of deviations from the probable in the case of a correlated system of variables is such that it can be reasonably supposed to have arisen from random sampling." Philosophical Magazine, 1900.  

Perkins, William, Mark Tygert, and Rachel Ward. "Computing the confidence levels for a root-mean-square test of goodness-of-fit." Applied Mathematics and Computation, 2011. https://doi.org/10.1016/j.amc.2011.03.124

Pettitt, Anthony N., Stephens, Michael A. "The Kolmogorov–Smirnov goodness-of-fit statistic with discrete and grouped data." Technometrics, 1977. https://doi.org/10.2307/1268631 

Phipson, Belinda, and Gordon K. Smyth. "Permutation p-values should never be zero: calculating exact p-values when permutations are randomly drawn." Statistical Applications in Genetics and Molecular Biology, 2010. https://doi.org/10.2202/1544-6115.1585

Ward, Rachel and Raymond J. Carroll. "Testing Hardy–Weinberg equilibrium with a simple root-mean-square statistic." Biostatistics, 2014. https://doi.org/10.1093/biostatistics/kxt028

Wood, Constance L., and Michele M. Altavela. “Large-Sample Results for Kolmogorov–Smirnov Statistics for Discrete Distributions.” Biometrika, 1978.
https://doi.org/10.1093/biomet/65.1.235
