---
title: "Introducing discretefit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{package_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package `discretefit` implements fast, Monte Carlo simulations for goodness-of-fit (GOF) tests for discrete distributions. This includes tests based on the root-mean-square statistic, the Chi-squared statistic, the log-likelihood-ratio statistic (G^2), and the Kolmogovov-Smirnov statistic.

Simulations are written in C++ (utilizing `Rcpp`) and are much faster than the simulated Chi-squared GOF test in the R `stats` package and the simulated Kolmogorov-Smirnov GOF test in the `dgof` package. 

## Usage

The GOF tests in `discretefit` function on a vector of counts, x, and a vector of probabilities, p. In the below example, x represents a vector of counts for five categories, and p represents a vector of probabilities for each corresponding category. The GOF tests provides p-values for the null hypothesis that x is a random sample of the discrete distribution defined by p. 

```{r setup, warning = FALSE}
library(discretefit)
library(bench)

x <- c(42, 0, 13, 2, 109)
p <- c(0.2, 0.05, 0.1, 0.05, 0.6)

pp <- c(rep(1, 4),
        rep(2, 1),
        rep(3, 2),
        rep(4, 1),
        rep(5, 12))

chisq_gof(x, p)
rms_gof(x, p)
ks_gof(x, p)
g_gof(x, p)
```

## Speed

The simulated Chi-squared GOF test in `discretefit` produces identical answers to the simulated Chi-squared GOF test in the stats package that is part of base R. 

```{r}
set.seed(499)
chisq_gof(x, p, reps = 2000)
set.seed(499)
chisq.test(x, p = p, simulate.p.value = TRUE)$p.value
```

However, because the Monte Carlo simulations in `discretefit` are written in C++, chisq_gof is much faster than `chisq.test`, especially when a large number of simulations are required.

```{r}
bench::system_time(
  chisq_gof(x, p, reps = 20000)
)

bench::system_time(
  chisq.test(x, p = p, simulate.p.value = TRUE, B = 20000)
)
```

The `ks_gof` function in `discretefit` is also faster than the simulated Kolmogorov-Smirnov test in the `dgof` package. (The `ks.test` function in the `stats` package in base R does not include a simulated test for discrete distributions.)

The p-values produced by `ks_gof`, however, are not exactly identical to those produced by ks.test in the `dgof` package because of slight variations in the algorithms. 

```{r}
x <- c(114, 118, 112, 158)
y <- c(1, 2, 3, 4, 5, 5)
p <- c(0.2, 0.2, 0.2, 0.4)
  
bench::system_time(
  ks_gof(x, p, reps = 20000)
)

bench::system_time(
  dgof::ks.test(x, ecdf(y), simulate.p.value = TRUE, B = 20000)
)
```

Additionally, the simulated GOF tests in base R and the `dgof` package are vectorized, so for large vectors attempting a large number of simulations maybe not be possible because of memory constraints. Since the function in `discretefit` are not vectorized, memory use is minimized. 

## Root-mean-square statistic

In a surprising number of cases, a simulated GOF test based on the root-mean-square statistic outperforms the Chi-squared test and other tests in the Cressie-Read power divergence family. This has been demonstrated by Perkins, Tygert, and Ward (2011). They provide the following toy example.

Take a discrete distribution with 50 bins (or categories). The probability for the first and second bin is 0.25. The probability for each of the remaining 48 bins is 0.5 / 48 (~0.0104).  

Now take the observed counts of 15 for the first bin, 5 for the second bin, and zero for each of the remaining 48 bins. It's obvious that these observations are very unlikely to be a random sample from the above distribution. However, the Chi-squared test and G^2 test fail to reject the null hypothesis. 

```{r}
x <- c(15, 5, rep(0, 48))
p <- c(0.25, 0.25, rep(1/(2 * 50 -4), 48))

chisq_gof(x, p)
g_gof(x, p)
```

By contrast, the root-mean-square test convincingly rejects the null hypothesis.

```{r}
rms_gof(x, p)
```

For additional examples, also see Perkins, Tygert, and Ward (2011) and Ward and Carroll (2014). 

## Alternatives 

As noted above, the `stats` package in base R implements a simulated  Chi-squared GOF test, and the `dgof` package implements simulated Kolmogorov-Smirnov GOF test.

I'm not aware of an R package that implements a simulated G^2 GOF test but the packages `RVAideMemoire` and `DescTools` implement GOF tests that utilize approximations based on the Chi-squared distribution. 

I'm not aware of another R package that implements a root-mean-square GOF test.

## References

Eddelbuettel, Dirk and Romain Francois. "Rcpp: Seamless R and C++ Integration." Journal of Statistical Software, 2011. https://www.jstatsoft.org/article/view/v040i08

Perkins, William, Mark Tygert, and Rachel Ward. "Computing the confidence levels for a root-mean-square test of goodness-of-fit." Applied Mathematics and Computation, 2011. https://doi.org/10.1016/j.amc.2011.03.124

Ward, Rachel and Raymond J. Carroll. "Testing Hardy–Weinberg equilibrium with a simple root-mean-square statistic." Biostatistics, 2014. https://doi.org/10.1093/biostatistics/kxt028
